# Introduction

\
This chapter 
definition, data literacy and proficiency, data structure, how data can be characterized.Prequisites R

## What is data?

Data can be understood as a collection of factual elements -- such as numbers, text, or recorded observations -- that represent information about the world. They:

-   can originate from various sources (e.g., public data such as census records, social media, web scraping, ...)\
-   can have various formats (e.g., quantitative, qualitative, ...)
-   can be processed and analyzed in order to gather knowledge and/or guide decision making

As data's volume, complexity and importance is growing and digitalization and AI are increasingly shaping our everyday lives (not only across academic disciplines, but also throughout the private sector), **solid data skills are among the key skills of our time!**

![](figures/intro/data3){width="60%" fig-align="center"}

## What are data literacy and data proficiency?

Data literacy and data proficiency are two levels of data capability that enable people to work with data in a meaningful way. There is no unique definition of data literacy and data proficiency in the literature. Data proficiency is sometimes considered a subfield of data literacy or the terms are used interchangeably. The definitions presented below are based on a Canadian study of the Dalhousie University [@ridsdale2015strategies] that was picked up by the German Informatics Society (GI) [@heidrich2018data] and on an article by @wagner25.

**Data literacy** is a set of fundamental skills, that focus on basic understanding, interpreting, and communicating data in context. It comprises:

-   *Conceptual understanding:* understanding different data sources and data collection methods, as well as their inherent limitations; understanding different data types and basic statistical concepts of appropriate evaluation and presentation methods; interpretation of statistics and visualization methods (i.e., recognizing patterns and trends in the data).

-   *Basic technical skills:* knowledge of basic data organization and data (pre-)processing methods and tools.

-   *Critical awareness:* ability to identify non-reliable data sources and data quality issues, and to recognize obvious flaws in data collection, analysis, presentation, and interpretation; it also involves critical questioning of conclusions drawn from data based on the presented evidence, as well as basic knowledge of data ethics.

-   *Communication:* presenting (aggregated) data in an appropriate way and translating results into everyday language that can also be understood by non-specialists; it also involves understanding of data-based explanations provided by others.

**Data proficiency** (sometimes also referred to as **data fluency**) is an advanced level of data capability that builds on data literacy. It focuses on data collection and processing, statistical analysis, and informed decision-making. It comprises:

-   *Methodological understanding:* thorough understanding of statistical concepts and methods that enables users to select appropriate (advanced) statistical methods based on the specific research question; the ability to identify complex patterns and relationships that might remain hidden when using simpler analysis techniques and the ability to identify methodological flaws that might undermine the validity of conclusions drawn from data analyses.

-   *Data analysis skills:* development of data analysis plans and conduct of statistical analyses.

-   *Advanced technical skills:* application of data organization and data (pre-)processing methods and tools; use of software for data science tasks, such as R and Python.

-   *Cross-functional communication:* translating research questions into appropriate data analysis tasks; translating complex statistical findings into action plans.

## From structured to unstructured data

Depending on the requirements and restrictions of the data, one distinguishes between the following three formats:

**Structured data:** This is the most common data type in simple use cases. Structured data are organized in a fixed format, such as rows and columns, and are easier to manipulate and analyze with traditional tools than more complex data types. A drawback is, that they are not very flexible as one has to stick to the given format. Examples are:

-   Spreadsheets (e.g., Excel, Google sheets)

    ![Source: <https://r4ds.hadley.nz/spreadsheets.html>](figures/intro/excel){width="65%" fig-align="center"}

-   Text files (e.g., CSV)

-   Relational data (e.g., SQL)

    ![](figures/intro/sql){width="65%" fig-align="center"}

**Unstructured data:** Unstructured data are not organized in a predefined manner. They often come in complex forms such as text documents, images, and videos. A key advantage is that they offer full flexibility as one is not restricted to a specific format. However, their analysis requires more sophisticated methods, like natural language processing (NLP) and machine learning techniques, as they doesn't easily fit into traditional relational databases. Unstructured data often play a key role in sentiment analysis and complex pattern recognition. Examples are texts and social media (instagram, twitter, etc.).

![](figures/intro/unstructured){width="65%" fig-align="center"}

**Semi-structured data:** Semi-structured data cannot be (directly) stored in a relational database, but have some organizational properties. Often a processed versions of semi-structured data can be stored therein. They are more flexible than structured data, but less flexible than unstructured data. They are widely used in scenarios such as web scraping and data integration. Example are markup languages (such as HTML or XML, or even TeX) and JSON.

![Source: <https://help.seagullscientific.com/2022/en/Content/DB_Types_XML.htm>](figures/intro/XML){width="65%" fig-align="center"}

## Characterizing data {#sec-data}

The aim of any data analysis is usually to draw conclusions about a specific population of interest. We refer to:

-   *population:* entire group of individuals or items that you want to draw conclusions about

-   *sample:* a subset of the population (oftentimes randomly drawn)

    ![Source: <https://www.questionpro.com/blog/population-vs-sample/>](figures/intro/sample){width="55%" fig-align="center"}

-   *unit of analysis:* the primary entity or object of study (e.g., person, object, geographical area,...)

-   *variable / feature / endpoint:* a characteristic of that unit of analysis that is being measured (e.g., age, income, pace,...)

It is often not possible to collect data from the entire population, so a smaller sample is studied instead. Depending on the size, quality and representativeness of the sample, it can be used to infer characteristics of the full population (see @sec-inferential).

Every variable has a data type (i.e., scale of measurement) that determines what kind of data (e.g., text, numbers,...) it can hold. The data type is crucial for choosing the correct data analysis method. We distinguish between the following data types:

**Quantitative data:** Quantitative data represent information that can be counted or measured in numbers. They can be further divided in **discrete** (i.e., countable) data (e.g., number of cigarettes smoked per day, number of cars caught speeding per day) and **continuous** (i.e., uncountable) data (e.g., speed, gas price, body weight) data.

**Qualitative data:** Qualitative data is non-numerical information. They can be further divided in **ordinal** (i.e., ordered categories) data (e.g., ratings on a likert scale, grades, education level) and **nominal** (i.e., uncountable) data (e.g., gender, hair color, nationality) data.

![Source: <https://www.vecteezy.com/vector-art/30821802-4-types-of-data-with-nominal-ordinal-discrete-and-continuous-data>](figures/intro/datatypes2){fig-align="center"}

Note that continuous variables are sometimes measured on a discrete scale (e.g., age in years). Nevertheless, we will usually analyze them as continuous variables. Likewise, discrete variables with many possible outcomes are sometimes analyzed as continuous variables, whereas quantitative variables might be analyzed like categorical ones, when only a few different outcomes are observed. Any continuous variable can be transformed into a categorical one by dividing its outcomes in discrete intervals (discretization). However, this inherently results in a loss of information.

A further characteristic of data is the **data dimennsionality**. We distinguish:

-   *no dimensional* data: no dimensionality available (e.g., graphs or strings)

-   *univariate* feature $X$: one value $x_i$ per observation $i$

-   *bivariate* feature $(X_1, X_2)$: two values $(x_{i1}, x_{i2})$ per observation $i$

-   *multivariate* feature $(X_1,\ldots,X_p)$: multiple values $(x_{i1},\ldots,x_{ip})$ per observation $i$

-   *high-dimensional* data: multivariate feature with large $p$


## Programming
Programming is a tool that you will need if you are not only interested in understanding the theoretical foundation of data literacy and data proficiency, but also want to practically perform data analyses. \
Throughout the course we will use the software R to demonstrate practical application of the methods presented.
R is a free, open-source programming language and software environment designed for statistical computing, data analysis, and high-quality graphics. It can handle a wide range of basic and complex analyses making it a valuable asset for many data-related tasks.

This chapter provides basic instructions on how to get started with R. For a more detailed introduction to the R language, please refer to [Hands on Programming](<https://rstudio-education.github.io/hopr/index.html>) by @grolemund14. 

### Installing R and RStudio
Use this link [<https://cloud.r-project.org/>] to download R. Select the proper version for your laptop.

![](figures/intro/R){fig-align="center"}

RStudio is an integrated development environment that provides a user-friendly interface for writing and managing R code. You can download and install it from [<https://posit.co/downloads/>].

![](figures/intro/RStudio){fig-align="center"}

After you have installed both R and RStudio, you can start using R by opening the RStudio program. We recommend to update R and RStudio regularly. Sometimes this requires you to re-install all your packages.

### Installing R packages
R packages are collections of R functions, data, and compiled code that extend the functionality of the base R language. They are typically located in an online  repository that is accessible to everyone. Some of the most popular repositories for R packages are:

- [CRAN](https://cran.r-project.org/): The Comprehensive R Archive Network is the official repository and maintained by the R community around the world. It features $> 20,000$ available packages. Package that are  published here need to adhere to CRAN policies and must pass several tests. 

- [Bioconductor](https://www.bioconductor.org/): This is a repository focusing on the development and sharing of open source software for analysis of biological data. Like CRAN, it has its own submission and review processes.

- [Github](https://github.com/): A proprietary developer platform (not R-specific) providing unlimited public and private repositories, robust documentation and support, encouragement of collaboration on code, content, and research. There is no strict review process associated with it.

- [kaggle](https://www.kaggle.com/datasets): A data science competition platform and online community providing powerful tools and resources for data analysis.


Packages can be installed using the `install.packages()` function. The following line of code installs the tidyverse package. Type it in the console, then press enter to run it.
```{r}
#| label: install
#| eval: false

install.packages("tidyverse")
```

Once installed, a package has to be loaded into the session to be used.
```{r}
#| label: load
#| eval: false

library(tidyverse)
```

To install more than one R package at a time, just write them as a character vector in the first argument of the `install.packages()` function.
```{r}
#| label: sev.pack
#| eval: false

install.packages("MASS", "rms")
```

To unload a given package, you can use the `detach()` function.
```{r}
#| label: unload
#| eval: false

detach("package:tidyverse", unload=TRUE)
```


### Sources of documentation and help

Obtain general overview of the package.
```{r}
#| label: help
#| eval: false

library(MASS) ## load the package
help(package = "MASS")
```

Obtain documentation for a specific function from the package.
```{r}
#| label: help2
#| eval: false

help(boxcox, package = "MASS")
```

For some packages, there is additional documentation in the form of vignettes. These are documents in which the authors present some of the features of their package in more detail. 
The list of vignettes existing for an installed packages can be obtained via the function `browseVignettes()`. A browser window will open and you can click on the preferred vignette to open it.
```{r}
#| label: help3
#| eval: false
 
library(tidyverse) ## load the package
browseVignettes(package="tidyverse")
```

If you prefer to stay in the command line, use the `vignette()` function to get a list of existing vignettes or to obtain the vignette you want to explore.
```{r}
#| label: vignette
#| eval: false

vignette(package = "tidyverse")
vignette("manifesto") 
```



